---
title: "phytoTaxonomy"
author: "B"
date: "April 7, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(stringsAsFactors = FALSE) 
library(knitr)
library(tidyverse)
library(readxl)
library(R.utils)  #function "capitalize"
library(taxize)
library(gdata)

read1<-function(dir='data/raw_data/',file='nla2012_waterchem_wide.csv'){
  a<-tryCatch(read.csv(paste(dir,file,sep=''),sep=',',na.strings = c(""," ","NA","N/A")),warning = function(e) read.csv(paste('../',dir,file,sep=''),sep=',',na.strings = c(""," ","NA","N/A")))
  return(a)
}  
```

## Introduction

* The goal is to get a harmonized list of phytoplankton taxa for the 2007 and 2012 National Lakes Assessments.

## Data Steps

###phtyoSp2007
* read the data
* change all occurences of '(Undetermined)' to NA
* change all SPECIES=="sp." to NA
* delete observations where GENUS==NA or odd values (e.g,"Encapsulating Chrysophyte" )
    - some species like 
* Change genus name "Isthmochoron" to "Isthmochloron" in fields "OTU", "TAXANAME", and "GENUS_CNT"
* Change genus name to "Sphaerodinium" for GENUS=="Sphaerodinium/Glenodinium/Peridiniopsis Complex"
* Change genus name to "Staurodesmus" for GENUS=="Staurodesmus/Arthrodesmus/Octacanthium"
* combine species names into a single field "sp2007"
* In the past we deleted the following taxa; keep them?????
    - *Chamaesiphon*
    - *Cuspidothrix issatschenkoi* NOTE: Homotypic Synonyms are: *Anabaena issatschenkoi* and *Aphanizomenon issatschenkoi*
    - *Eucocconeis*
    - *Psammothidium*
    - *Stigeoclonium*
* change all names to lowercase
* chg(2007) 'taxatype' to 'algal_group'
* chg(2007) 'division' to 'phylum'
* write table 'output/phytoSp2007.csv'

```{r phyto2007, include=FALSE, echo=FALSE}
p2007<-read1('data/raw_data/','nla2007_phyto_count.csv')

#change all occurences of '(Undetermined)' to NA
p2007[p2007=='(Undetermined)']<-NA

#remove non-specific genus names
p2007<-mutate(p2007,GENUS=ifelse(GENUS=="Encapsulating Chrysophyte",NA,GENUS))
p2007<-mutate(p2007,GENUS=ifelse(GENUS=="Dinobryon (empty lorica)",NA,GENUS))

#update taxonomy
p2007<-mutate(p2007,GENUS=ifelse(GENUS=="Sphaerodinium/Glenodinium/Peridiniopsis Complex","Sphaerodinium",GENUS))
#p2007<-mutate(p2007,GENUS=ifelse(GENUS=="Staurodesmus/Arthrodesmus/Octacanthium","Staurodesmus",GENUS))
#p2007<-mutate(p2007,OTU=ifelse(OTU=="Isthmochoron labulatum","Isthmochloron labulatum",OTU))
p2007<-mutate(p2007,TAXANAME=ifelse(TAXANAME=="Isthmochoron labulatum","Isthmochloron labulatum",TAXANAME))
p2007<-mutate(p2007,GENUS_CNT=ifelse(GENUS_CNT=="Isthmochoron","Isthmochloron",GENUS_CNT))


#create output file
phytoSp2007<-filter(p2007,!is.na(GENUS))%>%
            mutate(SPECIES=ifelse(SPECIES=="sp.",NA,SPECIES),
              species=trimws(ifelse(is.na(SPECIES),GENUS,ifelse(is.na(VARIETY),paste(GENUS,SPECIES),paste(GENUS,SPECIES,VARIETY)))))%>%
              select(TAXATYPE,DIVISION,ORDER,FAMILY,GENUS,species)%>%
              distinct()%>%arrange(DIVISION,GENUS)%>%
                group_by(GENUS) %>%
                mutate(rank=row_number(species))%>%
                spread(rank,species)%>%
                  mutate(species=ifelse(is.na(`1`),NA,paste(`1`,`2`,`3`,`4`,`5`,`6`,`7`,`8`,`9`,`10`,sep='; ')))%>%
                    mutate(species=gsub('NA; ','',species))%>%
                      mutate(species=gsub('; NA','',species))%>%
                        select(TAXATYPE,DIVISION,ORDER,FAMILY,GENUS,sp2007=species)%>%ungroup()

names(phytoSp2007)<-tolower(names(phytoSp2007))
phytoSp2007<-rename(phytoSp2007,phylum=division,algal_group=taxatype)

#View(phytoSp2007)
                        
#write.table(phytoSp2007,'output/phytoSp2007.csv',sep=',',row.names=FALSE)

```

###phtyoSp2012
* read the data
* delete observations where GENUS==NA
* combine species names into a single field "sp2012"
* change all taxa names to first letter capitalized to match 2007
* change all names to lowercase
* change SPECIES==*"EUASTRUM BIDNETATUM"* to SPECIES==*EUASTRUM BIDENTATUM*
* change SPECIES==*"PSEUDOKEPHYRION PSUEDOSPIRALE"* to SPECIES==*PSEUDOKEPHYRION PSEUDOSPIRALE*
* change SPECIES==*"PSEUDOKEPHYRION PSUEDOSPIRALE"* to SPECIES==*PSEUDOKEPHYRION PSEUDOSPIRALE*


toupper('Pseudokephyrion psuedospirale')

```{r phyto2012, include=FALSE, echo=FALSE}
p2012<-read1('data/raw_data/','nla2012_phyto_taxa.csv')

#spelling corrections
  p2012<-mutate(p2012,SPECIES=ifelse(SPECIES=='EUASTRUM BIDNETATUM','EUASTRUM BIDENTATUM',SPECIES))
  p2012<-mutate(p2012,SPECIES=ifelse(SPECIES=='PSEUDOKEPHYRION PSUEDOSPIRALE','PSEUDOKEPHYRION PSEUDOSPIRALE',SPECIES))
  p2012<-mutate(p2012,SPECIES=ifelse(SPECIES=='PSEUDOKEPHYRION PSUEDOSPIRALE','PSEUDOKEPHYRION PSEUDOSPIRALE',SPECIES))

nrow(p2012)  #885
nrow(select(p2012,TAXA_ID)%>%distinct()) #885

phytoSp2012<-filter(p2012,!is.na(GENUS))%>%
          mutate(species=ifelse(is.na(SPECIES),GENUS,SPECIES))%>%
              select(ALGAL_GROUP,PHYLUM,CLASS,ORDER,FAMILY,GENUS,species)%>%
              distinct()%>%arrange(PHYLUM,GENUS)%>%
                mutate_each(funs(tolower))%>%mutate_each(funs(capitalize))%>%
                group_by(GENUS) %>%
                mutate(rank=row_number(species))%>%
                spread(rank,species)%>%
                  mutate(species=ifelse(is.na(`1`),NA,paste(`1`,`2`,`3`,`4`,`5`,`6`,`7`,`8`,`9`,
                          `10`,`11`,`12`,`13`,`14`,`15`,`16`,`17`,`18`,`19`,`20`,`21`,`22`,`23`,`24`,sep='; ')))%>%
                    mutate(species=gsub('NA; ','',species))%>%
                      mutate(species=gsub('; NA','',species))%>%
                        select(ALGAL_GROUP,PHYLUM,CLASS,ORDER,FAMILY,GENUS,sp2012=species)%>%ungroup()
                          
names(phytoSp2012)<-tolower(names(phytoSp2012))

#View(phytoSp2012)
```

* create a master taxa list to compare
* get higher level classification from www.itis.gov
* join phytoSp2007, phytoSp2012 and itis to create phytoSp
* add genus_id as rownumber
* save itis file as 'phytoItis.csv'
* save phytoSp as 'phytoSp.csv'

```{r master, include=FALSE, echo=FALSE}
#join the 2007 and 2012 taxa lists
p07<-phytoSp2007
p07<-rename(p07,phylum07=phylum,algal_group07=algal_group,order07=order,family07=family)

phytoSp<-full_join(phytoSp2012,p07)

phytoSp<-select(phytoSp,algal_group12=algal_group,algal_group07,phylum12=phylum,phylum07,class12=class,order12=order,order07,family12=family,family07,genus,sp12=sp2012,sp07=sp2007)

write.csv(phytoSp,'output/phtyosp.csv')

#get higher classification from itis for each genus
# tsn<-classification(phytoSp$genus, db = 'itis',row=1)
# 
# itis<-c()
# for(i in c(1:length(tsn))){
#   if(is.na(tsn[[i]])==TRUE) {
#     itis<-bind_rows(itis,data.frame(genus=phytoSp$genus[i]))
#   } else {
#     x<-spread(tsn[[i]][-3],rank,name)
#     y<-spread(tsn[[i]][-1],rank,id)
#     names(x)<-paste(names(x),'itis',sep="_")
#     names(y)<-paste(names(y),'tsn',sep="_")
#     z<-data.frame(genus=phytoSp$genus[i])
#     itis<-bind_rows(itis,cbind(z,x,y))
#   }}
# 
# save(itis,file='data/workfiles/itis.rda')

#phytoSp<-select(phytoSp,algal_group12,algal_group07,phylum12,phylum07,phylum_itis,division_itis,class12,class_itis,order12,order07,order_itis,family12,family07,family_itis,genus,genus_itis,sp12,sp07)

#join the itis data
#load(file='data/workfiles/itis.rda')
#phytoSp<-full_join(phytoSp,itis)
#phytoSp$genus_id<-as.numeric(row.names(phytoSp))


#write.csv(itis,'output/phytoItis.csv',row.names=FALSE)
#write.csv(phytoSp,'output/phytoSp.csv',row.names=FALSE)
#itis<-read.csv('output/phytoItis.csv')
```

### get the algaebase classification by genus (thanks Betty)
* Betty meticulously scraped the taxonomy from algaebase.org for each taxon
* Get the 2007 and 2012 taxononmy
* join them and harmonize to the genus level


```{r betty, include=FALSE, echo=FALSE}
ab<-read.csv('data/workfiles/nla_algaebase_taxonomy.csv')
   names(ab)

nrow(ab) #362
nrow(phytoSp) #362

keep<-select(phytoSp,genus,sp07,sp12)

phytosp<-full_join(keep,ab)%>%select(phylum,subphylum,class,subclass,order,family,genus,genus_new,authority,sp07,sp12,flag,note)
nrow(phytosp) #362

write.csv(phytosp,'data/workfiles/nla_algaebase_taxonomy_to_check.csv',row.names = FALSE)


    
    
    
```
    
    