---
title: "nla2007_&_2012_data_for_farnaz_may2017"
author: "B"
date: "May 9, 2017"
output: html_document
---

## To Do
* `r paste("insult of the day:",bismer::insult())`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(stringsAsFactors = FALSE) 
library(knitr)
library(tidyverse)
library(readxl)

read1<-function(dir='data/raw_data/',file='nla2012_waterchem_wide.csv'){
  a<-tryCatch(read.csv(paste(dir,file,sep=''),sep=',',na.strings = c(""," ","NA","N/A")),warning = function(e) read.csv(paste('../',dir,file,sep=''),sep=',',na.strings = c(""," ","NA","N/A")))
  return(a)
}  

names1<-function(df) names(df)[order(names(df))]
```

## Introduction

* This document ([nla2007_&_2012_data_for_farnaz_may2017](https://github.com/willbmisled/lakes_database/blob/master/R/nla2007_&_2012_data_for_farnaz_may2017.Rmd)) describes how a subset of the 2007 and 2012 National Lake Assessment data were assembled and processed.
* The raw data are available on the NARS website:  https://www.epa.gov/national-aquatic-resource-surveys
* These data were processed and harmonized.  See the following documents for more information:
    - [nla2007_waterchem](https://github.com/willbmisled/lakes_database/blob/master/R/nla2007_waterchem.rmd)
    - [nla2012_waterchem](https://github.com/willbmisled/lakes_database/blob/master/R/nla2012_waterchem.rmd)
    - [nla_sites-and-nla_samples](https://github.com/willbmisled/lakes_database/blob/master/R/nla_sites-and-nla_samples.Rmd)
* The output includes the following datasets:
    - [farnaz201705.csv](https://github.com/willbmisled/lakes_database/blob/master/output/views/farnaz201705.csv)
    - [farnaz201705_meta.csv](https://github.com/willbmisled/lakes_database/blob/master/output/views/farnaz201705_meta.csv)

## Data Notes
* Decide whether or not to keep duplicate samples (...$duplicate=='D')
* Decide whether or not to keep samples from ...$visit_no=2 

## Data Steps
* read in the raw data and data definitions (see: https://github.com/willbmisled/lakes_database/tree/master/output)
    - 'nla2007_chem.csv'
    - 'nla2012_chem.csv'
    - 'nla_chem_data_defintions.csv'
    - 'nla_chem_parameters.csv'
    - 'nla_sites.csv'
    - 'nla_sites_meta.csv'
    - 'nla_samples.csv'
    - 'nla_samples_meta.csv'
* spread chem2007 and chem2012 based on uid, duplicate, parameter, and result then row_bind
* merge sites, sample, and chem data
* select fields to include and reorder

```{r data, include=FALSE, echo=FALSE}
#read the raw data files
chem2007<-read1('output/','nla2007_chem.csv')
chem2012<-read1('output/','nla2012_chem.csv')
dd<-read1('output/','nla_chem_data_defintions.csv')
pd<-read1('output/','nla_chem_parameters.csv')
sites<-read1('output/','nla_sites.csv')
sites_m<-read1('output/','nla_sites_meta.csv')
samples<-read1('output/','nla_samples.csv')
samples_m<-read1('output/','nla_samples_meta.csv')

#merge sites and samples
nrow(samples) #2476
nrow(sites) #2281
ss<-inner_join(samples,sites)
nrow(ss) #2476

#stack chem2007 and chem2012
s1<-select(chem2007,uid,duplicate,parameter,result)%>%
  spread(parameter,result)
nrow(s1) #1276
s1$year<-2007

s2<-select(chem2012,uid,duplicate,parameter,result)%>%
  spread(parameter,result)
nrow(s2) #1230
s2$year<-2012

chem<-bind_rows(s1,s2)

#merge chem and ss
nrow(chem) #2506
out<-left_join(chem,ss)

#select fields and reorger
out<-select(out,year,uid,site_id,visit_no,depthmax,lakearea,area_ha,lake_origin,fw_eco9,nut_reg,lat_dd83,lon_dd83,xcoord,ycoord,ammonia_n,chla,do2_2m,doc,microcystin,nh4ion,nitrate_n,nitrate_nitrite_n,nitrite_n,ntl,ph,ph_field,ptl,saxitoxin,secchi,silica,sulfate,toc,tss,turb)

```

```{r meta, include=FALSE, echo=FALSE}
#make sure the names match
names(dd)
names(pd)[1]<-"column_name"
    names(pd)
names(sites_m)
names(samples_m)
names(chem)

#select the column_names to include
dd1<-filter(dd,column_name%in%names(out)[names(out)%in%dd$column_name])
dd2<-filter(pd,column_name%in%names(out)[names(out)%in%pd$column_name])
dd3<-filter(sites_m,column_name%in%names(out)[names(out)%in%sites_m$column_name])
dd4<-filter(samples_m,column_name%in%names(out)[names(out)%in%samples_m$column_name])

#add year
dd5<-data.frame(column_name='year',description="sample year (e.g. 2007 or 2012)")

#bind them all
meta<-bind_rows(dd2,dd1,dd3,dd4,dd5)
meta<-arrange(meta,column_name)%>%distinct()

#check all are included
table(names(out)%in%meta$column_name)
```

* The final dataset and the data definitions saved as:
    - "farnaz201705.csv"
    - "farnaz201705_meta.csv"

```{r save, include=FALSE, echo=FALSE,eval=TRUE,eval=TRUE}
write.table(out,'../output/views/farnaz201705.csv',sep=',',row.names=FALSE)
write.table(meta,'../output/views/farnaz201705_meta.csv',sep=',',row.names=FALSE)

```





