---
title: "nla2012_raw_data"
author: "Bryan"
date: "February 23, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(stringsAsFactors = FALSE) 
library(knitr)

```

## Introduction
* This document describes the process of downloading and storing the raw data for the 2012 National Lakes Assessment.  For more information see [nla2012_master.rmd](https://github.com/willbmisled/lakes_database/tree/master/data/nla2012_master).

* The NLA2012 data are available online: https://www.epa.gov/national-aquatic-resource-surveys/data-national-aquatic-resource-surveys

* The NLA2012 data comprise 20 data and 19 metadata files (according to the website two of the datasets share metadata).  

* Some of the original metadata files are in tab delimited format. These were resaved as CSV files to match the other data and metadata files.

* Information on the raw data including the file type (data or metadata), the indicator (shortname for the dataset from the NARS website), dataset (our shorter name for the dataset), the final filename, the original filename (filename_old), and the URL is contained in the csv file (/data/raw_data/nla2012_raw_data_sources).

* The indicator, dataset, filename, and filename_old information for the data are shown below.


```{r kable, include=TRUE, echo=FALSE}
raw<-read.csv('../data/raw_data/nla2012_raw_data_sources.csv',sep=',')
kable(raw[,c(2:5)])
```

* The code used to download are in the two chunks below.

```{r get_data, include=TRUE, eval=FALSE}
a1<-read.csv('https://www.epa.gov/sites/production/files/2016-11/nla2012_algaltoxins_08192016.csv')
a2<-read.table('https://www.epa.gov/sites/production/files/2016-11/nla2012_algaltoxins_meta_08192016.txt',sep='\t', header=TRUE)
a3<-read.csv('https://www.epa.gov/sites/production/files/2016-11/nla2012_atrazine_08192016.csv')
a4<-read.table('https://www.epa.gov/sites/production/files/2016-11/nla2012_atrazine_meta_08192016.txt',sep='\t', header=TRUE)
a5<-read.csv('https://www.epa.gov/sites/production/files/2016-11/nla2012_bentcond_08232016.csv')
a6<-read.table('https://www.epa.gov/sites/production/files/2016-11/nla2012_bentcond_meta_08232016.txt',sep='\t', header=TRUE)
a7<-read.csv('https://www.epa.gov/sites/production/files/2016-11/nla2012_bentmet.csv')
a8<-read.table('https://www.epa.gov/sites/production/files/2016-11/nla2012_bentmet_meta.txt',sep='\t', header=TRUE)
a9<-read.csv('https://www.epa.gov/sites/production/files/2016-11/nla2012_benttaxa_wide_10272015.csv')
a10<-read.table('https://www.epa.gov/sites/production/files/2016-11/nla2012_benttaxa_wide_meta_10272015.txt',sep='\t', header=TRUE)
a11<-read.csv('https://www.epa.gov/sites/production/files/2016-11/nla2012_chla_wide.csv')
a12<-read.table('https://www.epa.gov/sites/production/files/2016-11/nla2012_chla_wide_meta.txt',sep='\t', header=TRUE)
a13<-read.csv('https://www.epa.gov/sites/production/files/2016-12/nla_2012_condition_categories.csv')
a14<-read.csv('https://www.epa.gov/sites/production/files/2016-12/nla_2012_condition_categories_meta.csv')
a15<-read.csv('https://www.epa.gov/sites/production/files/2016-12/nla2012_phytocnt_02122014.csv')
a17<-read.csv('https://www.epa.gov/sites/production/files/2016-12/nla2012_phytotaxa_wide_10272015.csv')
a18<-read.table('https://www.epa.gov/sites/production/files/2016-12/nla2012_phytotaxa_wide_meta_10272015.txt',sep='\t', header=TRUE)
a19<-read.csv('https://www.epa.gov/sites/production/files/2016-12/nla2012_secchi_08232016.csv')
a20<-read.table('https://www.epa.gov/sites/production/files/2016-12/nla2012_secchi_meta_08232016.txt',sep='\t', header=TRUE)
a21<-read.csv('https://www.epa.gov/sites/production/files/2016-12/nla2012_topsedhg_08192016.csv')
a22<-read.table('https://www.epa.gov/sites/production/files/2016-12/nla2012_topsedhg_meta_08192016.txt',sep='\t', header=TRUE)
a23<-read.csv('https://www.epa.gov/sites/production/files/2016-12/nla2012_waterchem_wide.csv')
a24<-read.table('https://www.epa.gov/sites/production/files/2016-12/nla2012_waterchem_meta.txt',sep='\t', header=TRUE)
a25<-read.csv('https://www.epa.gov/sites/production/files/2016-11/nla2012_wide_benthic_08232016.csv')
a26<-read.table('https://www.epa.gov/sites/production/files/2016-11/nla2012_wide_benthic_meta_08232016.txt',sep='\t', header=TRUE)
a27<-read.csv('https://www.epa.gov/sites/production/files/2016-12/nla2012_wide_phab_08232016_0.csv')
a28<-read.table('https://www.epa.gov/sites/production/files/2016-12/nla2012_wide_phab_meta_08232016.txt',sep='\t', header=TRUE)
a29<-read.csv('https://www.epa.gov/sites/production/files/2016-12/nla2012_wide_phabmet_10202016.csv')
a30<-read.table('https://www.epa.gov/sites/production/files/2016-12/nla2012_wide_phabmet_meta_08232016.txt',sep='\t', header=TRUE)
a31<-read.csv('https://www.epa.gov/sites/production/files/2016-12/nla2012_wide_profile_08232016.csv')
a32<-read.table('https://www.epa.gov/sites/production/files/2016-12/nla2012_wide_profile_meta_08232016.txt',sep='\t', header=TRUE)
a33<-read.csv('https://www.epa.gov/sites/production/files/2016-12/nla2012_wide_siteinfo_08232016.csv')
a34<-read.table('https://www.epa.gov/sites/production/files/2016-12/nla2012_wide_siteinfo_meta_08232016.txt',sep='\t', header=TRUE)
a35<-read.csv('https://www.epa.gov/sites/production/files/2016-12/nla2012_zoopcond_08192016.csv')
a36<-read.table('https://www.epa.gov/sites/production/files/2016-12/nla2012_zoopcond_meta_08192016.txt',sep='\t', header=TRUE)
a37<-read.csv('https://www.epa.gov/sites/production/files/2016-12/nla2012_zoopmets_08192016.csv')
a38<-read.table('https://www.epa.gov/sites/production/files/2016-12/nla2012_zoopmets_meta_08192016.txt',sep='\t', header=TRUE)
a39<-read.csv('https://www.epa.gov/sites/production/files/2016-12/nla2012_zooptaxa_wide_10272015.csv')
a40<-read.table('https://www.epa.gov/sites/production/files/2016-12/nla2012_zooptaxa_wide_meta_10272015.txt',sep='\t', header=TRUE)
```

```{r save_data, include=TRUE, eval=FALSE}
write.table(a1,'../data/raw_data/nla2012_algaltoxins_08192016.csv',row.names=F,sep=',')
write.table(a2,'../data/raw_data/nla2012_algaltoxins_meta_08192016.csv',row.names=F,sep=',')
write.table(a3,'../data/raw_data/nla2012_atrazine_08192016.csv',row.names=F,sep=',')
write.table(a4,'../data/raw_data/nla2012_atrazine_meta_08192016.csv',row.names=F,sep=',')
write.table(a5,'../data/raw_data/nla2012_bentcond_08232016.csv',row.names=F,sep=',')
write.table(a6,'../data/raw_data/nla2012_bentcond_meta_08232016.csv',row.names=F,sep=',')
write.table(a7,'../data/raw_data/nla2012_bentmet.csv',row.names=F,sep=',')
write.table(a8,'../data/raw_data/nla2012_bentmet_meta.csv',row.names=F,sep=',')
write.table(a9,'../data/raw_data/nla2012_benttaxa_wide_10272015.csv',row.names=F,sep=',')
write.table(a10,'../data/raw_data/nla2012_benttaxa_wide_meta_10272015.csv',row.names=F,sep=',')
write.table(a11,'../data/raw_data/nla2012_chla_wide.csv',row.names=F,sep=',')
write.table(a12,'../data/raw_data/nla2012_chla_wide_meta.csv',row.names=F,sep=',')
write.table(a13,'../data/raw_data/nla_2012_condition_categories.csv',row.names=F,sep=',')
write.table(a14,'../data/raw_data/nla_2012_condition_categories_meta.csv',row.names=F,sep=',')
write.table(a15,'../data/raw_data/nla2012_phytocnt_02122014.csv',row.names=F,sep=',')
write.table(a17,'../data/raw_data/nla2012_phytotaxa_wide_10272015.csv',row.names=F,sep=',')
write.table(a18,'../data/raw_data/nla2012_phytotaxa_wide_meta_10272015.csv',row.names=F,sep=',')
write.table(a19,'../data/raw_data/nla2012_secchi_08232016.csv',row.names=F,sep=',')
write.table(a20,'../data/raw_data/nla2012_secchi_meta_08232016.csv',row.names=F,sep=',')
write.table(a21,'../data/raw_data/nla2012_topsedhg_08192016.csv',row.names=F,sep=',')
write.table(a22,'../data/raw_data/nla2012_topsedhg_meta_08192016.csv',row.names=F,sep=',')
write.table(a23,'../data/raw_data/nla2012_waterchem_wide.csv',row.names=F,sep=',')
write.table(a24,'../data/raw_data/nla2012_waterchem_meta.csv',row.names=F,sep=',')
write.table(a25,'../data/raw_data/nla2012_wide_benthic_08232016.csv',row.names=F,sep=',')
write.table(a26,'../data/raw_data/nla2012_wide_benthic_meta_08232016.csv',row.names=F,sep=',')
write.table(a27,'../data/raw_data/nla2012_wide_phab_08232016_0.csv',row.names=F,sep=',')
write.table(a28,'../data/raw_data/nla2012_wide_phab_meta_08232016.csv',row.names=F,sep=',')
write.table(a29,'../data/raw_data/nla2012_wide_phabmet_10202016.csv',row.names=F,sep=',')
write.table(a30,'../data/raw_data/nla2012_wide_phabmet_meta_08232016.csv',row.names=F,sep=',')
write.table(a31,'../data/raw_data/nla2012_wide_profile_08232016.csv',row.names=F,sep=',')
write.table(a32,'../data/raw_data/nla2012_wide_profile_meta_08232016.csv',row.names=F,sep=',')
write.table(a33,'../data/raw_data/nla2012_wide_siteinfo_08232016.csv',row.names=F,sep=',')
write.table(a34,'../data/raw_data/nla2012_wide_siteinfo_meta_08232016.csv',row.names=F,sep=',')
write.table(a35,'../data/raw_data/nla2012_zoopcond_08192016.csv',row.names=F,sep=',')
write.table(a36,'../data/raw_data/nla2012_zoopcond_meta_08192016.csv',row.names=F,sep=',')
write.table(a37,'../data/raw_data/nla2012_zoopmets_08192016.csv',row.names=F,sep=',')
write.table(a38,'../data/raw_data/nla2012_zoopmets_meta_08192016.csv',row.names=F,sep=',')
write.table(a39,'../data/raw_data/nla2012_zooptaxa_wide_10272015.csv',row.names=F,sep=',')
write.table(a40,'../data/raw_data/nla2012_zooptaxa_wide_meta_10272015.csv',row.names=F,sep=',')
```