---
title: "nla2012_waterchem"
author: "Bryan Milstead"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(stringsAsFactors = FALSE) 
library(knitr)
library(tidyverse)
library(readxl)

read1<-function(dir='data/raw_data/',file='nla2012_waterchem_wide.csv'){
  a<-tryCatch(read.csv(paste(dir,file,sep=''),sep=',',na.strings = c(""," ","NA","N/A")),warning = function(e) read.csv(paste('../',dir,file,sep=''),sep=',',na.strings = c(""," ","NA","N/A")))
  return(a)
}  
```

```{r formatData_function, include=FALSE}
#function to reformat the raw NLA data based on a modified data structure file
formatData<-function(x='nla2012_waterchem.csv',y='waterchem_nla2012.csv'){
#read the raw data; trycatch will look for the data in two directories
a<-tryCatch(read.csv(paste('data/raw_data/',x,sep=''),sep=',',na.strings = c(""," ","NA","N/A")),
            warning = function(e) read.csv(paste('../data/raw_data/',x,sep=''),sep=',',na.strings = c(""," ","NA","N/A")))

#get the modified data structure
struc<-tryCatch(read.csv(paste('data/workfiles/',y,sep=''),sep=',',na.strings = c(""," ","NA","N/A")),
            warning = function(e) read.csv(paste('../data/workfiles/',y,sep=''),sep=',',na.strings = c(""," ","NA","N/A")))

#choose columns to keep, gather, or delete
keeps<-filter(struc,parameter=='column_name')%>%select(DATA_COLUMN)%>% unlist(use.names = FALSE)
gathers<-filter(struc,parameter!='column_name',column_name!='delete')%>%select(DATA_COLUMN)%>%distinct()%>% unlist(use.names = FALSE)
deletes<-filter(struc,parameter=='delete')%>%select(DATA_COLUMN)%>% unlist(use.names = FALSE)

#gather 
wc<-gather(a,DATA_COLUMN,value,one_of(gathers))

#deconstruct the COLUMN_NAME (parameter)
  #df "struc" has the new row and column names
    #delete unwanted columns
a<-left_join(wc,struc)%>%select(-one_of(deletes))

#create new df based on descontructed column_name
v<-distinct(a,column_name)%>% unlist(use.names = FALSE)
ch<-select(a,one_of(keeps),parameter) %>% distinct()  #dim(waterchem) 1326 27

for(i in c(1:length(v))){
  w<-filter(a,column_name==v[i])%>%
    select(UID,parameter,value)
  names(w)[3]<-v[i]
  ch<-full_join(ch,w)
}

#change column names to snake case
names(ch)<-tolower(names(ch))

#add source
ch$source<-x

#output
return(list(ch,struc))
}
```


## To Do
* change character to numeric for result, mdl, rl ???
* change col_names and character data to snake case
* `r paste("insult of the day:",bismer::insult())`

## Naming Conventions
* All variable names in lowercase
* use snake case for compound names (e.g., big_data)
* use parameter for the variable (e.g., NTL)
* use value for the measured value of the parameter

## Introduction

* This document ([nla2012_waterchem](https://github.com/willbmisled/lakes_database/blob/master/R/nla2012_waterchem.Rmd)) describes  converting the water chemistry data for the 2012 National Lakes Assessment from the wide format of the raw data files to a long format for database development.  For more information see [nla2012_master.rmd](https://github.com/willbmisled/lakes_database/blob/master/R/nla2012_master.Rmd).

* The output includes the following NLA 2012 datasets
    - waterchem
    - chla
    - algal toxins
    - atrazine
    - secchi
    - profile (partial: only the DO2_2M data are included)
    
* Four of these datasets (waterchem, chla, algal toxins, and atrazine) were processed in the same way:
    
    -  The original raw data files are in a wide format with two types of column names:
        - sample specific data: this includes things such as the UID (unique identifier for the sample), the LAB where the data were processed, and sample_ID.
        - parameter specific data: the names are complex variables indicating the a chemical parameter (e.g. COND=conductivity) and what is reported (e.g RESULT, UNITS, BATCH_ID, etc.).  
        
    - Not all of these data are useful to us but we will keep most of them just in case.  The final product may be pruned.
    
    - To support the conversion to a long format data structure files were created manually as csv files.  The structure files define which columns are "sample specic", which are "parameter specific" and which should be deleted.  They will be used to convert the paramater specific data to a long format while the sample specific data are kept in the wide format.  Confusing?

* The final two datasets secchi and profile deal with single parameters (secchi and DO2_2M). See details below
 

## Data Steps

* a function "formatData" was written to:
    - import the data and the structure file
    - reformat the data to a long format
* each dataset is reformatted and the general structure is defined
* datasets were combined into a single data.frame
* rearrange
* rename
* snake_case
    
### waterchem data

```{r waterchem, include=FALSE, echo=FALSE}
q<-formatData('nla2012_waterchem.csv','waterchem_nla2012.csv')
waterchem<-q[[1]]
struc_waterchem<-q[[2]]

#change nitrate_nitrite_n' units to 'mg N/L' (mostly correct but a few observation have units=='mg/l)'
waterchem<-mutate(waterchem,units=ifelse(parameter=='nitrate_nitrite_n',units=='mg N/L',units))
```

* The following table shows how the raw data will be deconstructed:
    - DATA_COLUMN: represents the original columns in the raw data
    - parameter:  in most cases this will be the parameter name in the new dataset except:
        - parameter=="column_name" indicates this column is sample specific so all unique observations will be kept
        - parameter=="delete" indicates this column will be deleted from the final data

```{r k_waterchem, include=TRUE, echo=FALSE}
kable(struc_waterchem)

```

* The waterchm dataset has the following columns (NOTE: some of these may be dropped or renamed in the final dataset):

```{r c_waterchem, include=TRUE, echo=FALSE}   
colnames(waterchem)
```

* The following parameters are included as rows in the final dataset (NOTE: some of these may be dropped or renamed in the final dataset):
    
```{r p_waterchem, include=TRUE, echo=FALSE}   
unique(waterchem$parameter)
```

### chla

* Chlorophyll was measured in the littoral zone (CHLL) and the deep spot (CHLX) these parameters were renamed:
    - CHLL = 'chla_littoral'
    - CHLX = 'chla'
    
```{r chla, include=FALSE, echo=FALSE}
q<-formatData('nla2012_chla.csv','chla_nla2012.csv')
chla<-q[[1]]
struc_chla<-q[[2]]
```

* The following table shows how the raw data will be deconstructed:
    - DATA_COLUMN: represents the original columns in the raw data
    - parameter:  in most cases this will be the parameter name in the new dataset except:
        - parameter=="column_name" indicates this column is sample specific so all unique observations will be kept
        - parameter=="delete" indicates this column will be deleted from the final data

```{r k_chla, include=TRUE, echo=FALSE}
kable(struc_chla)
```

* The chla dataset has the following columns (NOTE: some of these may be dropped or renamed in the final dataset):

```{r c_chla, include=TRUE, echo=FALSE}   
colnames(chla)
```

* The following parameters are included as rows in the final dataset (NOTE: some of these may be dropped or renamed in the final dataset):
    
```{r p_chla, include=TRUE, echo=FALSE}   
unique(chla$parameter)
```

### microcystin

* Microcystin was measured in the littoral zone (MICL) and the deep spot (MICX) these parameters will be renamed:
    - MICL = 'microcystin_littoral'
    - MICX = 'microcystin'
    
```{r mcl, include=FALSE, echo=FALSE}
q<-formatData('nla2012_algal_toxins.csv','algal_toxins_nla2012.csv')
mcl<-q[[1]]
struc_mcl<-q[[2]]
```

* The following table shows how the raw data will be deconstructed:
    - DATA_COLUMN: represents the original columns in the raw data
    - parameter:  in most cases this will be the parameter name in the new dataset except:
        - parameter=="column_name" indicates this column is sample specific so all unique observations will be kept
        - parameter=="delete" indicates this column will be deleted from the final data

```{r k_mcl, include=TRUE, echo=FALSE}
kable(struc_mcl)
```

* The mcl dataset has the following columns (NOTE: some of these may be dropped or renamed in the final dataset):

```{r c_mcl, include=TRUE, echo=FALSE}   
colnames(mcl)
```

* The following parameters are included as rows in the final dataset (NOTE: some of these may be dropped or renamed in the final dataset):
    
```{r p_mcl, include=TRUE, echo=FALSE}   
unique(mcl$parameter)
```

### atrazine

```{r atrazine, include=FALSE, echo=FALSE}
q<-formatData('nla2012_atrazine.csv','atrazine_nla2012.csv')
atrazine<-q[[1]]
struc_atrazine<-q[[2]]
```

* The following table shows how the raw data will be deconstructed:
    - DATA_COLUMN: represents the original columns in the raw data
    - parameter:  in most cases this will be the parameter name in the new dataset except:
        - parameter=="column_name" indicates this column is sample specific so all unique observations will be kept
        - parameter=="delete" indicates this column will be deleted from the final data

```{r k_atrazine, include=TRUE, echo=FALSE}
kable(struc_atrazine)
```

* The atrazine dataset has the following columns (NOTE: some of these may be dropped or renamed in the final dataset):

```{r c_atrazine, include=TRUE, echo=FALSE}   
colnames(atrazine)
```

* The following parameters are included as rows in the final dataset (NOTE: some of these may be dropped or renamed in the final dataset):
    
```{r p_atrazine, include=TRUE, echo=FALSE}   
unique(atrazine$parameter)
```

### secchi

```{r secchi, include=FALSE, echo=FALSE, eval=TRUE}
#read the raw data; trycatch will look for the data in two directories
secchi<-read1('data/raw_data/','nla2012_secchi.csv')

#add parameter and UNITS
secchi$parameter<-'secchi'
secchi$UNITS<-'m'

#rename and reorder
secchi<-select(secchi,uid=UID,parameter,result=SECCHI,units=UNITS,clear_to_bottom=CLEAR_TO_BOTTOM)

#clear to bottom
secchi$result[secchi$secchi_clear_to_bottom=='Y' & !is.na(secchi$clear_to_bottom)]<-NA

#add source
secchi$source<-'nla2012_secchi.csv'
```

* This dataset is mostly in the correct format.
* NOTE: changed all values for secchi to NA if CLEAR_TO_BOTTOM=='Y'
* Added: UNITS='m'
* The table below shows the columns that were added, deleted, or renamed.

```{r k_secchi, include=TRUE, echo=FALSE}
#get the modified data structure
struc_secchi<-tryCatch(read.csv('data/workfiles/secchi_nla2012.csv',sep=',',na.strings = c(""," ","NA","N/A")),warning = function(e) read.csv('../data/workfiles/secchi_nla2012.csv',sep=',',na.strings = c(""," ","NA","N/A")))
kable(struc_secchi)
```

### do2

```{r do2, include=FALSE, echo=FALSE, eval=TRUE}
#read the raw data; trycatch will look for the data in two directories
do2<-read1('data/raw_data/','nla2012_profile.csv')

#many observations with UIDs are missing SITE_ID
table(is.na(do2$SITE_ID)) #F==13762 T==16955
nrow(filter(do2,is.na(SITE_ID))) #16955

#check if the UID's are in the info file
i12<-read1('data/raw_data/','nla2012_site_info.csv')

nrow(filter(do2, !UID%in%i12$UID))  #16955
nrow(filter(do2,is.na(SITE_ID))) #16955

#harvest the UID and DO2_2M fields; delete rows with is.na(do2$SITE_ID)
do2<-filter(do2,!is.na(DO2_2M),!is.na(SITE_ID))%>%select(UID,DO2_2M)
nrow(do2)  #1189
nrow(filter(do2, !UID%in%i12$UID))  #0


do2<-select(do2,UID,DO2_2M)%>%filter(!is.na(DO2_2M))
nrow(do2)  #1189

#add UNITS='mg/l'
do2$UNITS<-'mg/l'

#add parameter
do2$parameter<-'do2_2m'

#rename and reorder
do2<-select(do2,uid=UID,parameter,result=DO2_2M,units=UNITS)

#add source
do2$source<-'nla2012_profile.csv'


```

* There are 16955 rows in the do2 data that have UIDs but no SITE_IDs.  None of these UIDs is represented in the siteInfo so these were dropped.
* Only the UID and DO2_2M columns are harvested from the profile dataset.
* DO2_2M is their calculation of the average dissolved oxygen in the top 2 meters of the lake.  It is based on the profile data.
* UNITS='mg/l' was added 


### join the datasets

* convert (if exist) fields RESULT, MDL, RL, and LOQ to numeric
* all files joined
* delete rows for is.na(result) 
* add field duplicate=='P'; for the 2007 data there are primary and duplicate value; for consistency we'll add the same field to the 2012 data and set all values to P=primary.
* as a qa/qc check the sums of the results by parameters were calculated and compared to the column sums for the raw data; all matched.
* check that all uid are in 'nla2007_site_info.csv'
* check that all uid correspond to STATUS=="Target_Sampled"


```{r join, include=FALSE, echo=FALSE,eval=TRUE}
waterchem<-mutate(waterchem,result=as.numeric(result),mdl=as.numeric(mdl),rl=as.numeric(rl))
chla<-mutate(chla,result=as.numeric(result),mdl=as.numeric(mdl),rl=as.numeric(rl))
atrazine<-mutate(atrazine,result=as.numeric(result),loq=as.numeric(loq),mdl=as.numeric(mdl))
mcl<-mutate(mcl,result=as.numeric(result),mdl=as.numeric(mdl),rl=as.numeric(rl))

chem2012<-bind_rows(waterchem,chla,mcl,atrazine,secchi,do2)

chem2012$duplicate='P'

chem2012<-filter(chem2012,!is.na(result))

#check that all uid are in 'nla2012_site_info.csv' and that they correspond to LAKE_SAMP=="Target_Sampled"
i12<-select(read1('data/raw_data/','nla2012_site_info.csv'),SITE_ID,UID,STATUS) #get the siteinfo
names(i12)<-tolower(names(i12))
nrow(chem2012) #32810
a<-left_join(chem2012,i12)%>%filter(status!="Target_Sampled")%>%select(uid,site_id,status)%>%distinct()
nrow(chem2012) #32810  all uid in 'nla2012_site_info.csv'
```

###QA/QC

* two qa/qc checks were run to make sure the data were assembled correctly.  
- sums by parameter were calculated and compared for the chem2012 and 'nla2012_waterchem.csv' data; all matched
- sums by uid were calculated and compared for the chem2012 and 'nla2012_waterchem.csv' data; all matched
* did not check sums for other datasets but the waterchem check shows the process worked.

```{r qa, include=FALSE, echo=FALSE,eval=TRUE}
#get the raw data and change names to lowercase
a<-read1('data/raw_data/','nla2012_waterchem.csv') 

#keep results and uid
a<-select(a,1,grep('_RESULT',names(a)))

#rename to match chem2012
names(a)<-tolower(unlist(strsplit(names(a),"_RESULT")))

#read the parameters from chem2012
param<-filter(chem2012,source=='nla2012_waterchem.csv')%>%select(parameter)%>%distinct()
param<-param$parameter

#restrict raw data to params and uid in chem2012
table(a$uid%in%chem2012$uid) #T=1230
table(chem2012$uid%in%a$uid) #T=32810

#get sums by parameter for raw and chem2012
rawParam<-apply(a[,-1],2,sum,na.rm=TRUE)
rawParam<-data.frame(parameter=names(rawParam),rawSum=rawParam)

chemParam<-group_by(chem2012, parameter) %>%
  summarize(chemSum=sum(result, na.rm = TRUE))

test<-inner_join(chemParam,rawParam)%>%mutate(test=chemSum==rawSum)
all.equal(test$chemSum,test$rawSum)

#compare  sums by UID
#get sums by parameter for raw and chem2012
names(a[,-1])
unique(chem2012$parameter)

b<-filter(chem2012,parameter%in%names(a[,-c(1:2)]))

table(b$parameter%in%names(a[,-c(1:2)]))
table(names(a[,-c(1:2)])%in%b$parameter)

rawUid<-data.frame(uid=a$uid,sumRaw=apply(a[,-c(1:2)],1,sum,na.rm=TRUE))

chemUid<-group_by(b, uid) %>%
  summarize(sumChem=sum(result, na.rm = TRUE))

test1<-inner_join(chemUid,rawUid)%>%mutate(test=sumChem==sumRaw)
all.equal(test1$sumChem,test1$sumRaw)

```

## Data definitions for the NLA2007 and NLA2012 chemistry data

* Below are the data definitions for column names of the chemistry data
* Included are four fields
    - **name**: is the column name (field)
    - **2007**: 1=field included in chem2007
    - **2012**: 1=field included in chem2012
    - **definition**: a brief description of the field 

```{r data_defintions, include=TRUE, echo=FALSE}
dd<-read1('output/','nla2007_2012_chem_data_defintions.csv')
kable(dd)
```

* The following table defines the parameters in the chem data
* This table has six fields
    - **parameter**: the name of the chemical analyte or measurement 
    - **name07**: the original name for the parameter in the NLA2007 data (if measured)
    - **name12**: the original name for the parameter in the NLA2012 data (if measured)
    - **units07**: the measurement units for the parameter in the NLA2007 data 
    - **units12**: the measurement units for the parameter in the NLA2012 data 
    - **definition**: a brief description of the parameter 
    
```{r param_defintions, include=TRUE, echo=FALSE}
pd<-read1('output/','nla2007_2012_chem_parameters.csv')
kable(pd)
```
## Save the data

* chem2012 saved as 'output/nla2012_chem.csv'

```{r save, include=FALSE, echo=FALSE,eval=TRUE}
tryCatch(write.table(chem2012,'output/nla2012_chem.csv',sep=',',row.names=FALSE),warning = function(e) write.table(chem2012,'../output/nla2012_chem.csv',sep=',',row.names=FALSE))
```



