---
title: nla2012_metadata
author: Bryan
date: February 23, 2017
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(stringsAsFactors = FALSE) 
library(knitr)
library(tidyverse)
```
## To Do
* Check that the metadata for "phytocnt" are in the metadata document for "phytotaxa"
* convert all variable names in the metadata and field to snake_case
* no description for "SAM_CODE" in the metadata record for waterchem

## Naming Conventions
* All variable names in lowercase
* use snake case for compound names (e.g., big_data)

## Introduction

* This document ([nla2012_metadata.rmd](https://github.com/willbmisled/lakes_database/blob/master/R/nla2012_metadata.Rmd)) describes the process of downloading and compiling the metadata for the 2012 National Lakes Assessment.  For more information see [nla2012_master.rmd](https://github.com/willbmisled/lakes_database/blob/master/R/nla2012_master.Rmd).

* The NLA2012 data are available online: https://www.epa.gov/national-aquatic-resource-surveys/data-national-aquatic-resource-surveys

* The NLA2012 data comprise 20 data and 19 metadata files (according to the website two of the datasets share metadata).

* All files were downloaded to the directory ['data/raw_data'](https://github.com/willbmisled/lakes_database/tree/master/data/raw_data)

* Information on the download  and saving process are in [nla2012_raw_data.rmd](https://github.com/willbmisled/lakes_database/blob/master/R/nla2012_raw_data.Rmd)

* The metadata for the 2012 NLA only includes the data dictionaries for the 20 data files.  At some point we will need to generate complete metadata records.

* Each dataset is a flat file that contains the data collected for each indicator group and some general information on the sampling and each site.  The indicator data should be unique but the general information is repeated in every dataset.

* For our database we will combine all of the metadata into a single file and then generate two csv file to store it.  The two files are:
    - [nla2012_metadat_long.csv](https://github.com/willbmisled/lakes_database/blob/master/data/nla2012_metadat_long.csv): a file with all of the metadata records for each dataset together in long format (see below).  This will be useful for extracting the data definitions for each dataset.  It includes repeats of some data definitions.
    - [nla2012_metadat_wide.csv](https://github.com/willbmisled/lakes_database/blob/master/data/nla2012_metadat_wide.csv): this file has the data definitions in wide format; the rows are the unique data definitions and the columns indicate which datasets they belong to.
 
### Data steps   

* GOAL: to combine all the metadata files into a single file and to remove redundancies.

* Read and combine the raw metadata records

* First problem, column_names are not the same for all metadata files.  
    - Each record has a column with the variable name; The name of this column is not standardized.  Three names are used: "PARAMETER", "VARIABLE", or "DATA_COLUMN".  For all files the column name was changed to "field"
    - Most metadata files have a column called "LABEL" for each variable's data definition.  In some files, however, this information is in two columns; one called "LABEL" and the other "DESCRIPTION".  Fortunately for each line only one or the other is used. The data from both columns (LABEL and DESCRIPTION) were combined and named "description"

* Decide which field to keep. The fields to delete are:
    - "TABLE_NAME": Not included in all meta data records; this seems to be a shorthand for the datasource
    - "SAMPLE_TYPE": This seems to be a shorthand for the datasource; may relate to table name in the original database used to derive the flat files.
    - "LEGAL_VALUES": for categorical data these are the acceptable values.  Not consistent across datasets so it will be eliminated.  Refer to the original metadata if this is of interest.
    - "RANGE_HIGH": probably for QA. Appears to be the highest acceptable values for the field.  Not consistent across datasets so it will be eliminated.  Refer to the original metadata if this is of interest.
    - "RANGE_LOW": probably for QA. Appears to be the lowest acceptable values for the field.  Not consistent across datasets so it will be eliminated.  Refer to the original metadata if this is of interest.
    
* Fields to keep are  "dataset", "field", "units", and "description".
* There is some redundancy in the data.frame so select the unique values

```{r harvest, include=FALSE, echo=FALSE}
#get a list of the metadata files
m<-read.csv('../data/nla2012_raw_data_sources.csv',sep=',')

m<-filter(m,type=='metadata')[-8,] %>%
    select(dataset,filename)

meta<-c() #empty space to hold the data
for(i in c(1:nrow(m))){
#read the data
  a<-read.csv(paste('../data/raw_data/',m$filename[i],sep=''),sep=',',na.strings = c(""," ","NA","N/A"))
#rename some column_names (if they exist) 
  names(a)[names(a)%in%c("PARAMETER","VARIABLE")]<-"DATA_COLUMN" 
#delete field "TABLE_NAME"  
  a<-select(a, -matches("TABLE_NAME"))
#add filename
  a$dataset<-m$dataset[i]
#add to meta
  meta<-bind_rows(meta,a)
}


#combine DESCRIPTION and LABEL then drop LABEL
meta$DESCRIPTION<-ifelse(is.na(meta$LABEL),meta$DESCRIPTION,meta$LABEL)
meta<-select(meta, -matches("LABEL"))

#remove leading and trailing spaces from all fields
  meta$description<-trimws(meta$DESCRIPTION)
  meta$description<-trimws(meta$DATA_COLUMN)
  meta$description<-trimws(meta$UNITS)
  meta$description<-trimws(meta$dataset)

#rename 'DATA_COLUMN' to 'field'
#rename 'DESCRIPTION' to 'description'
#keep columns dataset, field, units, and description

#get unique values for all fields

nrow(meta) #1926

meta<-select(meta,dataset,field=DATA_COLUMN,units=UNITS,description=DESCRIPTION)%>%
  distinct()

nrow(meta) #1893
```

* Check the field and DESCRIPTION fields for redundancy
    - many of the description fields are used for multiple column_names; mostly in the phab data; fine as is.
    - some of the DATA_COLUMN have slight variations in the descriptions.  We will choose one description for each DATA_COLUMN value.
    - YCOORD has the wrong information.  We can recover the correct description from "XCOORD"

* Check the units field for inconsistencies and redundancy
    - Change "M" to "m"
    - Change "NA", "None", "NONE" to NA

```{r newDesc, include=FALSE, echo=FALSE}
#check lengths and look for redundancy in the column names and descriptions
length(meta$field) #1893
length(unique(meta$field)) #1716
length(unique(meta$description)) #1655

chk<-distinct(select(meta, field, description)) 
nrow(chk) #1726

des<-as.data.frame(table(chk$description))
rep<-filter(des,Freq>1)[,1]
temp<-arrange(filter(chk,description%in%rep),description)
#write.table(temp, file='../data/temp.csv',row.names=F,sep=',')

col<-as.data.frame(table(chk$field))
rep1<-filter(col,Freq>1)[,1]
temp1<-arrange(filter(chk,field%in%rep1),field)
#write.table(temp1, file='../data/temp1.csv',row.names=F,sep=',')

#update the descriptions to eliminate redundancy
newDesc<-data.frame(field=c('DEPTH','DUPLICATE_DO','FFG','LAB','ODOR','ODOR_DESC','REVIEWED_BY_INITIAL','TAXA_ID','TOTALHG_RESULT','UID','YCOORD'),new=c('Depth at collection point','Duplicate DO reading taken','Functional feeding group; codes separated by commas if more than one','Lab name','Odor present','Description of odor','Intials of form reviewer','Taxa identification number','Analyte value for sediment mercury for Total Mercury','Unique identifier for sampling event','y-coordinate from US Contiguous Albers Equal Area Conic projection EXCEPT Hawaii sites which are calculated from Hawaii Albers Equal Area Conic projection'))

a<-full_join(meta,newDesc)
a$description[!is.na(a$new)]<-a$new[!is.na(a$new)]
meta<-select(a,-new)

#check lengths and look for redundancy in the column names and descriptions
length(meta$field) #1893 unchanged
length(unique(meta$field)) #1716 unchanged
length(unique(meta$description)) #1655 change to 1646

nrow(distinct(select(meta, field, description))) #1726 change to 1716

#check for inconsistencies in the field "units"
arrange(distinct(select(meta, units)),units)

#make some changes for consistency
  #replace "M" with "m" (meters)
meta<-mutate(meta,units=ifelse(units=='M','m',units))

meta<-mutate(meta,units=ifelse(units=='None' | units=='NONE' | units=='NA',NA,units))

#check for multiple values per field
chk<-distinct(select(meta, field, units)) 
nrow(chk) #1716
des<-as.data.frame(table(chk$field))
rep<-filter(des,Freq>1)[,1]  #none
```

* 
* Create the meta_wide field-a crosstab showing the fields(rows) and the datasets (columns)
    - still some redundancy in the long form.  Change all crosstab values > 1 to 1
* Create the final output files in csv format
    - nla2012_metadat_long.csv
    - nla2012_metadat_wide.csv
* Write CSV files to the ['data'](https://github.com/willbmisled/lakes_database/tree/master/data/raw_data) directory

```{r spread, include=FALSE, echo=FALSE, warning=FALSE}
#create pivot table of field(rows) by dataset(columns)
a<-spread(as.data.frame(table(meta$field,meta$dataset)),Var2,Freq)%>%
    rename(field=Var1)

#join values of units and description 
#get unique vales for field, units, and description
b<-distinct(select(meta, field, units,description))
#join to pivot
meta_wide<-left_join(b,a)

#change values >1 to 1
meta_wide[,4:22][meta_wide[,4:22]>1]<-1


#sort the data.frames
meta<-arrange(meta,field)
meta_wide<-arrange(meta_wide,field)

#write the final files
write.table(meta, file='../data/nla2012_metadat_long.csv',row.names=F,sep=',')
write.table(meta_wide, file='../data/nla2012_metadat_wide.csv',row.names=F,sep=',')
```

### Data definitions for the metadata files

* [nla2012_metadat_long.csv](https://github.com/willbmisled/lakes_database/blob/master/data/nla2012_metadat_long.csv) has `r nrow(meta)` rows and `r ncol(meta)` columns. The data definitions for the column names (fields) are: 

field  | description
------------- | -------------------------------------------------------------------------
dataset  | shorthand name of original file.  See: [nla2012_raw_data_sources.csv](https://github.com/willbmisled/lakes_database/blob/master/data/nla2012_raw_data_sources.csv)
field | the variable name
units | if applicable, the units for the variable
description | a description of the variable

* [nla2012_metadat_wide.csv](https://github.com/willbmisled/lakes_database/blob/master/data/nla2012_metadat_wide.csv) has `r nrow(meta_wide)` rows and `r ncol(meta_wide)` columns. The data definitions for the column names (fields) are: 

field  | description
------------- | -------------------------------------------------------------------------
field | the variable name
units | if applicable, the units for the variable
description | a description of the variable
algaltoxins | Reference to dataset in [nla2012_metadat_wide.csv](https://github.com/willbmisled/lakes_database/blob/master/data/nla2012_metadat_wide.csv)); 1 = this variable included in the dataset; 0 = not included. 
atrazine | Reference to dataset in [nla2012_metadat_wide.csv](https://github.com/willbmisled/lakes_database/blob/master/data/nla2012_metadat_wide.csv)); 1 = this variable included in the dataset; 0 = not included.     
bentcond | Reference to dataset in [nla2012_metadat_wide.csv](https://github.com/willbmisled/lakes_database/blob/master/data/nla2012_metadat_wide.csv)); 1 = this variable included in the dataset; 0 = not included.     
benthic  | Reference to dataset in [nla2012_metadat_wide.csv](https://github.com/willbmisled/lakes_database/blob/master/data/nla2012_metadat_wide.csv)); 1 = this variable included in the dataset; 0 = not included.     
bentmet  | Reference to dataset in [nla2012_metadat_wide.csv](https://github.com/willbmisled/lakes_database/blob/master/data/nla2012_metadat_wide.csv)); 1 = this variable included in the dataset; 0 = not included.    
benttaxa | Reference to dataset in [nla2012_metadat_wide.csv](https://github.com/willbmisled/lakes_database/blob/master/data/nla2012_metadat_wide.csv)); 1 = this variable included in the dataset; 0 = not included.     
chla | Reference to dataset in [nla2012_metadat_wide.csv](https://github.com/willbmisled/lakes_database/blob/master/data/nla2012_metadat_wide.csv)); 1 = this variable included in the dataset; 0 = not included.         
condition | Reference to dataset in [nla2012_metadat_wide.csv](https://github.com/willbmisled/lakes_database/blob/master/data/nla2012_metadat_wide.csv)); 1 = this variable included in the dataset; 0 = not included.    
phab | Reference to dataset in [nla2012_metadat_wide.csv](https://github.com/willbmisled/lakes_database/blob/master/data/nla2012_metadat_wide.csv)); 1 = this variable included in the dataset; 0 = not included.         
phabmet | Reference to dataset in [nla2012_metadat_wide.csv](https://github.com/willbmisled/lakes_database/blob/master/data/nla2012_metadat_wide.csv)); 1 = this variable included in the dataset; 0 = not included.      
phytotaxa | Reference to dataset in [nla2012_metadat_wide.csv](https://github.com/willbmisled/lakes_database/blob/master/data/nla2012_metadat_wide.csv)); 1 = this variable included in the dataset; 0 = not included.    
profile | Reference to dataset in [nla2012_metadat_wide.csv](https://github.com/willbmisled/lakes_database/blob/master/data/nla2012_metadat_wide.csv)); 1 = this variable included in the dataset; 0 = not included.      
secchi | Reference to dataset in [nla2012_metadat_wide.csv](https://github.com/willbmisled/lakes_database/blob/master/data/nla2012_metadat_wide.csv)); 1 = this variable included in the dataset; 0 = not included.      
siteinfo | Reference to dataset in [nla2012_metadat_wide.csv](https://github.com/willbmisled/lakes_database/blob/master/data/nla2012_metadat_wide.csv)); 1 = this variable included in the dataset; 0 = not included.     
topsedhg | Reference to dataset in [nla2012_metadat_wide.csv](https://github.com/willbmisled/lakes_database/blob/master/data/nla2012_metadat_wide.csv)); 1 = this variable included in the dataset; 0 = not included.     
waterchem | Reference to dataset in [nla2012_metadat_wide.csv](https://github.com/willbmisled/lakes_database/blob/master/data/nla2012_metadat_wide.csv)); 1 = this variable included in the dataset; 0 = not included.    
zoopcond | Reference to dataset in [nla2012_metadat_wide.csv](https://github.com/willbmisled/lakes_database/blob/master/data/nla2012_metadat_wide.csv)); 1 = this variable included in the dataset; 0 = not included.     
zoopmets | Reference to dataset in [nla2012_metadat_wide.csv](https://github.com/willbmisled/lakes_database/blob/master/data/nla2012_metadat_wide.csv)); 1 = this variable included in the dataset; 0 = not included.     
zooptaxa | Reference to dataset in [nla2012_metadat_wide.csv](https://github.com/willbmisled/lakes_database/blob/master/data/nla2012_metadat_wide.csv)); 1 = this variable included in the dataset; 0 = not included.   

