---
title: "nla2007_waterchem"
author: "Bryan Milstead"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(stringsAsFactors = FALSE) 
library(knitr)
library(tidyverse)
library(readxl)

read1<-function(dir='data/raw_data/',file='nla2012_waterchem_wide.csv'){
  a<-tryCatch(read.csv(paste(dir,file,sep=''),sep=',',na.strings = c(""," ","NA","N/A")),warning = function(e) read.csv(paste('../',dir,file,sep=''),sep=',',na.strings = c(""," ","NA","N/A")))
  return(a)
}    

names1<-function(df) names(df)[order(names(df))]

```

## To Do
* `r paste("insult of the day:",bismer::insult())`

## Naming Conventions
* All variable names in lowercase
* use snake case for compound names (e.g., big_data)
* use parameter for the variable (e.g., NTL)
* use value for the measured value of the parameter

## Introduction

* This document ([nla2007_waterchem](https://github.com/willbmisled/lakes_database/blob/master/R/nla_2007_waterchem.Rmd)) describes  converting the water chemistry data for the 2007 National Lakes Assessment from the wide format of the raw data files to a long format for database development.  For more information see [nla2012_master.rmd](https://github.com/willbmisled/lakes_database/blob/master/R/nla2012_master.Rmd).

* The output includes the following NLA 2007 datasets
    - waterchem (with chla & secchi)
    - algal toxins
    - profile (partial: only the DO2_2M data are included)
    
* Four of these datasets (waterchem, chla, algal toxins, and atrazine) were processed in the same way:
    
    -  The original raw data files are in a wide format with two types of column names:
        - sample specific data: this includes things such as the UID (unique identifier for the sample), the LAB where the data were processed, and sample_ID.
        - parameter specific data: the names are complex variables indicating the a chemical parameter (e.g. COND=conductivity) and what is reported (e.g RESULT, UNITS, BATCH_ID, etc.).  
        
    - Not all of these data are useful to us but we will keep most of them just in case.  The final product may be pruned.
    
    - To support the conversion to a long format data structure files were created manually as csv files.  The structure files define which columns are "sample specic", which are "parameter specific" and which should be deleted.  They will be used to convert the paramater specific data to a long format while the sample specific data are kept in the wide format.  Confusing?

* The final two datasets secchi and profile deal with single parameters (secchi and DO2_2M). See details below
 

## Data Steps

* a function "formatData" from nla2012_waterchem was re-written to:
    - import the waterchem data and the structure file
    - reformat the data to a long format

* datasets were combined into a single data.frame
* rearrange
* rename
* snake_case
    
##NLA 2007

###Waterchem

* Manually changed all column names in NLA2007_waterchem.csv for NA??? to SODIUM??? e.g. NA=SODIUM & NA_RL_ALERT=SODIUM_RL_ALERT
* The 2012 data use "UID" as a unique identifier for each sampling event
* For the 2007 data a unique event is defined by:
    - SITE_ID = unique id for the site
    - VISIT_NO = 1 or 2; all sites visited at least once (VISIT_NO == 1) but some were revisted (VISIT_NO == 2)
    - SAMPLE_CATEGORY = "P" or "D"; a primary sample ("P") collected from all sites and a field duplicate ("D") from some
* VISIT_ID appears to be unique to SITE_ID and VISIT_NO combinations.
    - in the dataset some "uid" are repeated because of duplicate samples (SAMPLE_CATEGORY=="D")
    - it is tempting to eliminate the duplicates but will keep them for now
* use VISIT_ID as the unique identifier ("uid")
* make sure all VISIT_ID are in 'nla_samples.csv' (TRUE)
* rename uid=VISIT_ID and add the prefix "2007_"
* restrict data to LAKE_SAMP=="Target_Sampled"
    - The following uid / duplicate combinations were deleted for LAKE_SAMP!="Target_Sampled"

```{r waterchem2007, include=FALSE}
#get the data
a<-read1('data/raw_data/','nla2007_waterchem.csv')
nrow(a) #1326

#get the sample info file
samples<-read1('output/','nla_samples.csv')%>%select(uid,site_id,visit_no)%>%filter(substr(uid,1,4)=="2007")

#restrict data to LAKE_SAMP=="Target_Sampled"
deleted_samples<-filter(a,LAKE_SAMP!="Target_Sampled"|is.na(LAKE_SAMP))%>%select(uid=VISIT_ID,site_id=SITE_ID,visit_no=VISIT_NO,duplicate=SAMPLE_CATEGORY,status=LAKE_SAMP)%>%mutate(uid=paste("2007_",uid,sep=''))
  
a<-filter(a,LAKE_SAMP=="Target_Sampled")
nrow(a) #1297



#check uid
uid<-mutate(a,uid=paste("2007_",VISIT_ID,sep=''))%>%select(uid,site_id=SITE_ID,visit_no=VISIT_NO)%>%distinct()
nrow(samples)#1246
nrow(uid) #1246
table(uid$uid%in%samples$uid) #T==1246
table(samples$uid%in%uid$uid) #T==1246

#get the modified data structure
struc<-read1('data/workfiles/','waterchem_nla2007.csv') 

#choose columns to keep, gather, or delete
keeps<-filter(struc,parameter=='column_name')%>%select(DATA_COLUMN)%>% unlist(use.names = FALSE)
gathers<-filter(struc,parameter!='column_name',column_name!='delete')%>%select(DATA_COLUMN)%>%distinct()%>% unlist(use.names = FALSE)
deletes<-filter(struc,parameter=='delete')%>%select(DATA_COLUMN)%>% unlist(use.names = FALSE)

#gather 
wc<-gather(a,DATA_COLUMN,value,one_of(gathers)) #dim 143967 49  NOTE: length(gathers)==111; nrow(a)==1297; 111*1297==143967

#deconstruct the COLUMN_NAME (parameter)
  #df "struc" has the new row and column names
    #delete unwanted columns
a<-left_join(wc,struc)%>%select(-one_of(deletes)) #dim(a) 143967     34

#table(a$parameter,a$column_name)

#create new df based on descontructed column_name
v<-distinct(a,column_name)%>% unlist(use.names = FALSE)
waterchem<-select(a,one_of(keeps),parameter) %>% distinct()  #dim(waterchem) 42801 27

for(i in c(1:length(v))){
  w<-filter(a,column_name==v[i])%>%
    select(SITE_ID,VISIT_NO,SAMPLE_CATEGORY,parameter,value)
  names(w)[5]<-v[i]
  waterchem<-full_join(waterchem,w)
}

#dim(waterchem) #42801    33

#change column names to snake case
names(waterchem)<-tolower(names(waterchem))

#add source
waterchem$source<-'nla2007_waterchem.csv'

# rename uid=VISIT_ID and add the prefix "2007_"
waterchem<-mutate(waterchem,uid=paste("2007_",visit_id,sep=''))%>%select(-visit_id)

```

```{r deleted, include=TRUE, echo=FALSE}
deleted_samples
```

### waterchem updates

* replace sample_depth with ph_field_depth for ph_field
* replace sample_id_chem with sample_id_chla for parameter=='chla'; rename sample_id_chem to sample_id
* replace lab_id_chem with lab_id_chla for parameter=='chla'; rename lab_id_chem to lab_id
* replace comment_fld_chem with comment_fld_chla for parameter=='chla'; rename comment_fld_chem to comment_fld

```{r wcUpdate, include=FALSE}
# replace sample_depth with ph_field_depth for ph_field
waterchem<-mutate(waterchem,sample_depth=ifelse(parameter=='ph_field',ph_field_depth,sample_depth))%>%
  select(-ph_field_depth)

# replace sample_id_chem with sample_id_chla for chla; rename sample_id_chem to sample_id
waterchem<-mutate(waterchem,sample_id=ifelse(parameter=='chla',sample_id_chla,sample_id_chem))%>%
  select(-sample_id_chla,-sample_id_chem)

# replace lab_id_chem with lab_id_chla for chla; rename lab_id_chem to lab_id
waterchem<-mutate(waterchem,lab_id=ifelse(parameter=='chla',lab_id_chla,lab_id_chem))%>%
  select(-lab_id_chla,-lab_id_chem)

# replace comment_fld_chem with comment_fld_chla for parameter=='chla'; rename comment_fld_chem to comment_fld
waterchem<-mutate(waterchem,comment_fld=ifelse(parameter=='chla',comment_fld_chla,comment_fld_chem))%>%
  select(-comment_fld_chla,-comment_fld_chem)

# replace comment_lab_chem with comment_lab_chla for parameter=='chla'; rename comment_lab_chem to comment_lab
waterchem<-mutate(waterchem,comment_lab=ifelse(parameter=='chla',comment_lab_chla,comment_lab_chem))%>%
  select(-comment_lab_chla,-comment_lab_chem)

```

### add units and rl to waterchem
* units harvested manually from the label field for "result" from the structure file
* rl (reporting limit) harvested manually from the label field for "rl_alert" from the structure file

```{r wcrl, include=FALSE}
new<-filter(struc,column_name=='result')%>%select(parameter,units,rl)
dim(waterchem) #42801    29

waterchem<-full_join(waterchem,new)
dim(waterchem) #42801    31

```

* The following table shows how the raw data will be deconstructed:
    - DATA_COLUMN: represents the original columns in the raw data
    - parameter:  in most cases this will be the parameter name in the new dataset except:
        - parameter=="column_name" indicates this column is sample specific so all unique observations will be kept
        - parameter=="delete" indicates this column will be deleted from the final data


```{r k_waterchem07, include=TRUE, echo=FALSE}
kable(struc)
```

* The waterchem dataset has the following columns (NOTE: some of these may be dropped or renamed in the final dataset):

```{r c_waterchem07, include=TRUE, echo=FALSE}   
names1(waterchem)
```

* The following parameters are included as rows in the waterchem dataset (NOTE: some of these may be dropped or renamed in the final dataset):
    
```{r p_waterchem07, include=TRUE, echo=FALSE}   
unique(waterchem$parameter)[order(unique(waterchem$parameter))]
```

###do2

* do2_2m is the average do2 reading for profile samples for depths less than 2m.
* we will calculate it directly from the profile data
* read the profile data
* add uid
* delete rows with uid==NA; these are the LAKE_SAMP!='Target_Sampled'


```{r do2, include=FALSE}
do2<-read1('data/raw_data/','nla2007_profile.csv')

#change column names to snake case
names(do2)<-tolower(names(do2))

#add the UID
nrow(uid) #1246
nrow(do2) #12670
do2<-(left_join(do2,uid))

#check the uids
nrow(do2) #12670
length(unique(do2$uid)) #1247
table(unique(do2$uid)%in%uid$uid) #1  1246 
filter(do2,!uid%in%uid$uid)%>%select(uid,site_id,visit_no)%>%distinct()
do2<-filter(do2,uid%in%uid$uid)
nrow(do2) #12659
length(unique(do2$uid)) #1246

#keep the profile data to use for temp below
  profile<-do2

# filter for depth<=2m
do2<-filter(do2,depth<=2)%>%select(uid,depth,do_field)

#calculate average do for depth<=2
do2 <- group_by(do2, uid)%>%
  summarise(result=mean(do_field, na.rm = TRUE))

#add UNITS='mg/l'
do2$units<-'mg/l'

#add parameter
do2$parameter<-'do2_2m'

#add source
do2$source<-'nla2007_profile.csv'

#add SAMPLE_CATEGORY
do2$sample_category<-'P'

#change column names to snake case
names(do2)<-tolower(names(do2))

```

###temperature   

* the field water temperature data are in the profile dataset as "TEMP_FIELD" for multiple depths
* depths are not standardized so we'll calculate
    - max_temp: max for the whole profile
    - min_temp: min for the whole profile
    - mean_temp: mean for the whole profile
    - mean_temp_2m: mean for the DEPTH<=2 meters
* rbind all temp variables
* add units, source, & sample_category
    
```{r temp, include=FALSE,eval=TRUE}

# max_temp
max_temp<- group_by(profile, uid)%>%
  summarise(result=max(temp_field, na.rm = TRUE))
    max_temp$parameter<-'temp_max'
      max_temp #1246

# min_temp     
min_temp<- group_by(profile, uid)%>%
  summarise(result=min(temp_field, na.rm = TRUE))
    min_temp$parameter<-'temp_min'
      min_temp #1246
      
# mean_temp
mean_temp<- group_by(profile, uid)%>%
  summarise(result=mean(temp_field, na.rm = TRUE))
    mean_temp$parameter<-'temp_mean'
      mean_temp #1246

# mean_temp_2m
      # filter for depth<=2m
profile2m<-filter(profile,depth<=2)%>%select(uid,temp_field)

#calculate average do for depth<=2
mean_temp_2m<- group_by(profile2m, uid)%>%
  summarise(result=mean(temp_field, na.rm = TRUE))
    mean_temp_2m$parameter<-'temp_2m'
      mean_temp_2m #1239
      
#rbind temp variables
field_temp<-rbind(min_temp,max_temp,mean_temp,mean_temp_2m)

#add UNITS='degrees C'
field_temp$units<-'degrees C'

#add source
field_temp$source<-'nla2007_profile.csv'

#add SAMPLE_CATEGORY
nrow(field_temp) #1252
length(unique(field_temp$uid)) #1252

field_temp$sample_category<-'P'

field_temp #4,977 (1246*3+1239) 
```

### algal toxins

* the algal toxins data were received directly from Keith Loftin at the USGS as an excel file.  This is saved in the directory 'data/raw_data/' as 'nla2007_algal_toxins.xlsx'.  
* the original file has a number of tables with ancillary data.  The algal toxins data are in a table named 'cyanotoxins ELISA'
* the cyanotoxins table was copied to a new spreadsheet and saved as 'nla2007_algal_toxins.csv'.
* columns with ancillary data (e.g. "State", "County", etc) were eliminated
* column names were changed to match the NLA data
* some data were modified
    - the column named "US EPA visit ID" contained a mashup of nla columns "visit_no" & sample_category; these were deconstructed to their original state
    - the column name "Total cylindrospermopsins conc ELISA         (µg/L)" was renamed "cylindrospermopsin"
    - the column name "Total microcystin conc ELISA (µg/L)" was renamed "microcystin"
    - the column name "Total saxitoxins conc ELISA (µg/L)" was renamed "saxitoxin"
* from the original column names we see that all units are in "ug/l"
* for values below the reporting limit are entered as character values (e.g. "<.05") these were all changed to zero; this also allowed us to harvest the following reporting limits:
    - cylindrospermopsin rl==.05
    - microcystin rl==.1
    - saxitoxin rl==.02
* for uid==7541 two samples were processed for the 3 algal toxins but both are listed as duplicate="P"
    - change one (usgs_id=="MM-09126A") to duplicate="D"

* here are Keith's data definitions for the algal toxin fields:
    - cylindrospermopsin: Total sample cylindrospermopsin concentration measured by ELISA in cylindrospermopsin equivalents (ug/L, micrograms per liter). Cylindrospermopsin concentration as measured by the Abraxis cylindrospermopsin enzyme-linked immunosorbentassay (ELISA) (PN: 522011) reported to represent the sum of all cross-reactive molecules from both intracellular and dissolved phases within a sample.  Samples were lysed by three sequential freeze/thaw cycles and filtered prior to analyses.
    - microcystin:  Total sample microcystin concentration measured by ELISA in microcystin-LR equivalents (ug/L, micrograms per liter). Microcystin/Nodularin concentration as measured by the Abraxis polyclonal "ADDA" enzyme-linked immunosorbentassay (ELISA) (PN: 520011) reported to represent the sum of all cross-reactive ADDA bearing molecules from both intracellular and dissolved phases within a sample.  Samples were lysed by three 
sequential freeze/thaw cycles and filtered before analyses.  The assay is capable of detecting both microcystins and nodularins, however, nodularins are considered to be predominately found in brackish waters.  As a result response for these results will be discussed in terms of microcystins, but in a rare cases nodularin has been measured in inland freshwaters.
    - saxitoxin: Total sample saxitoxin concentration measured by ELISA in saxitoxin equivalents (µg/L). Saxitoxin concentration as measured by the Abraxis saxitoxin (PSP) enzyme-linked immunosorbentassay (ELISA) (PN: 52255B) reported to represent the sum of all cross-reactive molecules from both intracellular and dissolved phases within a sample.  Samples were lysed by three sequential freeze/thaw cycles and filtered prior to analyses.
    
* data were read in 
* gathered into the long format
* units and rl added
* add uid
* delete rows with uid==NA; these are the LAKE_SAMP!='Target_Sampled'

```{r algal_toxins, include=FALSE, echo=FALSE,eval=TRUE, message=FALSE}
at<-read1('data/raw_data/','nla2007_algal_toxins.csv')

algal_toxins<-gather(at,key=parameter,value=result,cylindrospermopsin,microcystin,saxitoxin)

algal_toxins$units<-'ug/l'
algal_toxins<-mutate(algal_toxins,rl=ifelse(parameter=='cylindrospermopsin',.05,NA),
                                  rl=ifelse(parameter=='microcystin',.1,.02))

algal_toxins$source<-'nla2007_algal_toxins.csv'

#change sample_catergory from P to D for usgs_id=="MM-09126A"
algal_toxins<-mutate(algal_toxins,sample_category=ifelse(usgs_id=="MM-09126A","D",sample_category))

#add(uid)
algal_toxins<-mutate(algal_toxins,uid=paste("2007_",visit_id,sep=''))%>%select(-visit_id)

#check the uids
nrow(algal_toxins) #3981
length(unique(algal_toxins$uid)) #1252
table(unique(algal_toxins$uid)%in%uid$uid) #6  1246
filter(algal_toxins,!uid%in%uid$uid)%>%select(uid,site_id,visit_no)%>%distinct()
algal_toxins<-filter(algal_toxins,uid%in%uid$uid)
nrow(algal_toxins) #3963
```



### join the datasets and a few tweaks

* convert fields result and rl to numeric
* all files joined
* drop "site_id and visit_no"
* rename "sample_category" to "duplicate"
* check that all uid are in "nla_samples.csv" (TRUE)
* delete any uid that are in deleted_samples$uid (these were previously deleted for the waterchem data only)

```{r join, include=FALSE, echo=FALSE,eval=TRUE, message=FALSE}
#convert fields result and rl to numeric
waterchem<-mutate(waterchem,result=as.numeric(result),rl=as.numeric(rl))

#join files
chem2007<-bind_rows(waterchem,do2,field_temp,algal_toxins)

#drop "site_id and visit_no; rename "visit_ID" to "uid"; rename "sample_category" to "duplicate"
chem2007<-select(chem2007, -site_id,-visit_no)%>%rename(duplicate=sample_category)

#check uid
UID<-select(chem2007,uid)%>%distinct()
nrow(UID) #1246
table(UID$uid%in%samples$uid) #T=1246

#delete uid%in%deleted_samples$uid
chem2007<-filter(chem2007,!uid%in%deleted_samples$uid)
nrow(chem2007) #51968
```

###QA/QC

* Two qa/qc checks were run on the waterchem data to make sure the data were assembled correctly.
    - sums by parameter for the raw and final datasets were compared and were equal.
    - sums by uid for the raw and final datasets were compared and were equal.
* did not check sums for the algal_toxins or profile data but the waterchem check shows the process worked.

```{r qa, include=FALSE, echo=FALSE,eval=TRUE}
#get the raw data and change names to lowercase
a<-read1('data/raw_data/','nla2007_waterchem.csv') 

#restrict data to LAKE_SAMP=="Target_Sampled" & rename uid and duplicate
raw<-filter(a,LAKE_SAMP=="Target_Sampled")%>%mutate(uid=paste("2007_",VISIT_ID,sep=''),duplicate=SAMPLE_CATEGORY)

#compare chem sums between raw and final datasets
  #final sums are easy
final_sums<-group_by(chem2007, parameter)%>%summarise(sum=sum(result, na.rm = TRUE))
  #raw sums require some selection and name changes first
x<-filter(struc,column_name=='result')%>%select(parameter,DATA_COLUMN) #chem structure file with new and raw names
raw1<-select_(raw,.dots=x$DATA_COLUMN) #select results
names(raw1)<-x$parameter #rename
raw_sums<-apply(raw1,2,sum,na.rm=TRUE)
raw_sums<-data.frame(parameter=names(raw_sums),raw_sums=raw_sums)

compare<-full_join(final_sums,raw_sums)%>%mutate(test=sum==raw_sums)
table(compare$test,useNA='ifany') #TRUE==31 NA==8; the NA are non-Chem sums from chem2007

#compare  sums by UID
#get sums by parameter for raw and chem2007
raw_sums1<-apply(raw1,1,sum,na.rm=TRUE)
raw_sums1<-data.frame(uid=raw$uid,raw_sum=raw_sums1)
raw_sums1<-group_by(raw_sums1, uid)%>%summarise(raw_sum=sum(raw_sum, na.rm = TRUE))


final1<-filter(chem2007,parameter%in%x$parameter)
final_sums1<-group_by(final1, uid)%>%summarise(sum=sum(result, na.rm = TRUE))

compare1<-full_join(final_sums1,raw_sums1)%>%mutate(test=sum==raw_sum)
table(compare1$test,useNA='ifany') #TRUE==1228 NA==18; the falses are floating point errors
filter(compare1,test==FALSE)
```

## Data definitions for the NLA2007 and NLA2012 chemistry data

* check that all column names are in 'nla_chem_data_defintions.csv' (TRUE)
* Below are the data definitions for column names of the chemistry data
* Included are four fields
    - **column_name**: is the column name (field)
    - **name07**: the original name for the parameter in the NLA2007 data (if measured)
    - **name12**: the original name for the parameter in the NLA2012 data (if measured)
    - **description**: a brief description of the field
    
```{r check_meta, include=FALSE, echo=FALSE,eval=TRUE}
#check that all names are in the dd file
dd<-read1('output/','nla_chem_data_defintions.csv')
 table(names(chem2007)%in%dd$name)
 #names(chem2007)[!names(chem2007)%in%dd$name]  
 #filter(chem2007,!is.na("comment"))%>%select(source)%>%distinct()
 
#check that all parameters are in the pd file
pd<-read1('output/','nla_chem_parameters.csv')
table(unique(chem2007$parameter)%in%pd$parameter)
#filter(chem2012,!parameter%in%pd$parameter)%>%select(parameter,source)%>%distinct()
```

```{r data_defintions, include=TRUE, echo=FALSE,eval=TRUE}
kable(dd)
```

* check that all parameters are in 'nla_chem_data_paramters.csv' (TRUE)
* The following table defines the parameters in the chem data
* This table has six fields
    - **parameter**: the name of the chemical analyte or measurement 
    - **name07**: the original name for the parameter in the NLA2007 data (if measured)
    - **name12**: the original name for the parameter in the NLA2012 data (if measured)
    - **units07**: the measurement units for the parameter in the NLA2007 data 
    - **units12**: the measurement units for the parameter in the NLA2012 data 
    - **description**: a brief description of the parameter 
    
```{r param_defintions, include=TRUE, echo=FALSE,eval=TRUE}
kable(pd)
```

## spread check
* make sure the data can be "spread"; works

```{r spread, include=FALSE, echo=FALSE,eval=TRUE,eval=TRUE}
spread<-select(chem2007,uid,duplicate,parameter,result)%>%
  spread(parameter,result)

```

## Save the data

* chem2007 saved as 'output/nla2007_chem.csv'

```{r save, include=FALSE, echo=FALSE,eval=TRUE,eval=TRUE}
tryCatch(write.table(chem2007,'output/nla2007_chem.csv',sep=',',row.names=FALSE),warning = function(e) write.table(chem2007,'../output/nla2007_chem.csv',sep=',',row.names=FALSE))
```


## Data Notes

* Decide whether or not to keep duplicate samples (chem2007$duplicate=='D')
* To create a wide format verison of the data use: "chem2007_wide<-select(chem2007,uid,duplicate,parameter,result)%>%
  spread(parameter,result)"













