---
title: "nla2007_nla2012_site_info"
author: "B"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(stringsAsFactors = FALSE) 
library(knitr)
library(tidyverse)

read1<-function(dir='data/raw_data/',file='nla2012_waterchem_wide.csv'){
  a<-tryCatch(read.csv(paste(dir,file,sep=''),sep=',',na.strings = c(""," ","NA","N/A")),warning = function(e) read.csv(paste('../',dir,file,sep=''),sep=',',na.strings = c(""," ","NA","N/A")))
  return(a)
}    

```

## To Do
* create two files from the nla_site_info files
    - sample_info: the unique information about each sampling event
    - site_info: the unique information about each site
    - harmonize names between 2007 and 2012 whereever possible
* create data definitions file

* `r paste("insult of the day:",bismer::insult())`

## Naming Conventions
* All variable names in lowercase
* use snake case for compound names (e.g., big_data)
* use parameter for the variable (e.g., NTL)
* use value for the measured value of the parameter
* when names differ between the 2007 and 2012 files prefer the 2012 name; hopefully future NLA datasets will follow the 2012 naming conventions.

## Introduction

* This document shows how the NLA2007 and NLA2012 site_info files were processed.  The files contain the site and sampling event information along with the survey design info. Naming conventions were not conserved across years. Not all of the fields in these files are relevant to the uses of most researchers but will be kept for now.  Names will be harmonized across the datasets for the most important fields such as indices, and ancillary data.  As we find additional shared fields these names can be harmonized as well in later versions.

Each site_info file from the 2007 & 20012 NLA will be split into a site_info file and a sample_info file. The site_info file will have the design information and geographic data.  The sample_info file will have the data specific to each sampling event (e.g., date, what was collected, etc.).  The "site_id" will be used to link the two files.  The "uid" in the sample_info file will be used to link the sample_info to the nla data files (e.g. chem2007, chem2012, etc.)

## Data Steps

* read the two datafiles
    - i07<-'nla2007_site_info.csv'
    - i12<-'nla2012_site_info.csv'
    
* The 2007 site_info table for SITE_ID=="NLA06608-3846" is missing the VISIT_ID
    - add missing VISIT_ID==8844
    
* read the structure file (e.g. the combined data definitions file)
    - struc<-'nla2007_2012_site_and_sample_info_data_defintions.csv'

```{r getData, include=FALSE}
#get the data
i07<-read1('data/raw_data/','nla2007_site_info.csv')
i12<-read1('data/raw_data/','nla2012_site_info.csv')

#update missing VISIT_ID

filter(i07,SITE_ID=="NLA06608-3846")%>%select(SITE_ID,VISIT_ID)
i07<-mutate(i07,VISIT_ID=ifelse(SITE_ID=="NLA06608-3846",8844,VISIT_ID) )

#get the data structure file
struc<-read1('output/','nla2007_2012_site_and_sample_info_data_defintions.csv')
```
* below is the structure file.  This will also serve as the data definitions file for all four output files.  This file includes the following fields:
    - **site**: 1=include this field from the original site_info file in the output site_info file
    - **sample**: 1=include this field from the original site_info file in the output sample_info file
    - **column_name**: name of the field in the output file
    - **name07**: name of the field in the the original 2007 site_info file
    - **name12**: name of the field in the the original 2012 site_info file

```{r struc, include=TRUE, echo=FALSE}
kable(struc)
```

* split the site_info files into site_info and sample_info
* rename column based on the "struc" file
* NOTE: the site and sample files will only contain sites that were sampled.

```{r split, include=TRUE, echo=FALSE}
#site2007
site07<-filter(struc,site==1,!is.na(name07))%>%select(column_name,name07)
table(names(i07)%in%site07$name07)

site2007<-filter(i07,LAKE_SAMP=='Target_Sampled')%>%select_(.dots=site07$name07)%>%distinct()
all.equal(names(site2007),site07$name07)
names(site2007)<-site07$column_name

length(unique(site2007$site_id))  #1151 (of 1157)
nrow(site2007)  #1151 (of 1157)

#sample2007  
sample07<-filter(struc,sample==1,!is.na(name07))%>%select(column_name,name07)
table(names(i07)%in%sample07$name07)

sample2007<-filter(i07,LAKE_SAMP=='Target_Sampled')%>%select_(.dots=sample07$name07)
all.equal(names(sample2007),sample07$name07)
names(sample2007)<-sample07$column_name
nrow(sample2007) #1246

#site2012
site12<-filter(struc,site==1,!is.na(name12))%>%select(column_name,name12)
table(names(i12)%in%site12$name12)

site2012<-filter(i12,STATUS=='Target_Sampled')%>%select_(.dots=site12$name12)%>%distinct()
all.equal(names(site2012),site12$name12)
names(site2012)<-site12$column_name

length(unique(site2012$site_id))  #1130 of (2664)
nrow(site2012)  #1130 of (2664)

#sample2012  
sample12<-filter(struc,sample==1,!is.na(name12))%>%select(column_name,name12)
table(names(i12)%in%sample12$name12)

sample2012<-filter(i12,STATUS=='Target_Sampled')%>%select_(.dots=sample12$name12)
all.equal(names(sample2012),sample12$name12)
names(sample2012)<-sample12$column_name
nrow(sample2012) #1230
```

* Confirm that all chem2007 and chem2012 uid's are included.

```{r split, include=TRUE, echo=FALSE}
table(is.na(chem2007$uid)) #F==0  T==52962 
table(is.na(chem2012$uid)) #F==0  T==32848 

table(chem2007$uid%in%sample2007$uid,useNA='ifany') #F==0  T==52722 
table(chem2012$uid%in%sample2012$uid,useNA='ifany') #F==0  T==32810

```














