---
title: "phytoTaxonomy"
author: "B"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(stringsAsFactors = FALSE) 
library(knitr)
library(tidyverse)
library(readxl)
library(R.utils)  #function "capitalize"
library(taxize)
library(gdata)

read1<-function(dir='data/raw_data/',file='nla2012_waterchem_wide.csv'){
  a<-tryCatch(read.csv(paste(dir,file,sep=''),sep=',',na.strings = c(""," ","NA","N/A")),warning = function(e) read.csv(paste('../',dir,file,sep=''),sep=',',na.strings = c(""," ","NA","N/A")))
  return(a)
}  
```

## Introduction

* The goal is to get a harmonized list of phytoplankton taxa for the 2007 and 2012 National Lakes Assessments at the genus level.
* We started with a list of the genera and Betty scraped the current taxonomy from algaebase.org
* To this we will add a comma separted list of 2007 and 2012 species from the taxonomy lists.

## Data Steps

###phtyoSp2007
* read the data
* change all occurences of '(Undetermined)' to NA
* change all SPECIES=="sp." to NA
* delete observations where GENUS==NA or odd values (e.g,"Encapsulating Chrysophyte" )
* combine species names into a single field "sp2007"
* In the past we deleted the following taxa; keep them?????
    - *Chamaesiphon*
    - *Cuspidothrix issatschenkoi* NOTE: Homotypic Synonyms are: *Anabaena issatschenkoi* and *Aphanizomenon issatschenkoi*
    - *Eucocconeis*
    - *Psammothidium*
    - *Stigeoclonium*
* change all names to lowercase
* chg(2007) 'taxatype' to 'algal_group'
* chg(2007) 'division' to 'phylum'
* delete observations where GENUS==NA


```{r phyto2007, include=FALSE, echo=FALSE}
p2007<-read1('data/raw_data/','nla2007_phyto_count.csv')

#change all occurences of '(Undetermined)' to NA
p2007[p2007=='(Undetermined)']<-NA

#remove non-genus genus names
p2007<-mutate(p2007,GENUS=ifelse(GENUS=="Encapsulating Chrysophyte",NA,GENUS))
p2007<-mutate(p2007,GENUS=ifelse(GENUS=="Dinobryon (empty lorica)",NA,GENUS))

#create output file
phytoSp2007<-filter(p2007,!is.na(GENUS))%>%
            mutate(SPECIES=ifelse(SPECIES=="sp.",NA,SPECIES),
              species=trimws(ifelse(is.na(SPECIES),GENUS,ifelse(is.na(VARIETY),paste(GENUS,SPECIES),paste(GENUS,SPECIES,VARIETY)))))%>%
              select(TAXATYPE,DIVISION,ORDER,FAMILY,GENUS,species)%>%
              distinct()%>%arrange(DIVISION,GENUS)%>%
                group_by(GENUS) %>%
                mutate(rank=row_number(species))%>%
                spread(rank,species)%>%
                  mutate(species=ifelse(is.na(`1`),NA,paste(`1`,`2`,`3`,`4`,`5`,`6`,`7`,`8`,`9`,`10`,sep='; ')))%>%
                    mutate(species=gsub('NA; ','',species))%>%
                      mutate(species=gsub('; NA','',species))%>%
                        select(TAXATYPE,DIVISION,ORDER,FAMILY,GENUS,sp2007=species)%>%ungroup()

names(phytoSp2007)<-tolower(names(phytoSp2007))
phytoSp2007<-rename(phytoSp2007,phylum=division,algal_group=taxatype)

#View(phytoSp2007)
                        
#write.table(phytoSp2007,'output/phytoSp2007.csv',sep=',',row.names=FALSE)

```

###phtyoSp2012
* read the data
* delete observations where GENUS==NA
* combine species names into a single field "sp2012"
* change all taxa names to first letter capitalized to match 2007
* change all names to lowercase
* change SPECIES==*"EUASTRUM BIDNETATUM"* to SPECIES==*EUASTRUM BIDENTATUM*
* change SPECIES==*"PSEUDOKEPHYRION PSUEDOSPIRALE"* to SPECIES==*PSEUDOKEPHYRION PSEUDOSPIRALE*
* change SPECIES==*"PSEUDOKEPHYRION PSUEDOSPIRALE"* to SPECIES==*PSEUDOKEPHYRION PSEUDOSPIRALE*


```{r phyto2012, include=FALSE, echo=FALSE}
p2012<-read1('data/raw_data/','nla2012_phyto_taxa.csv')

#spelling corrections
  p2012<-mutate(p2012,SPECIES=ifelse(SPECIES=='EUASTRUM BIDNETATUM','EUASTRUM BIDENTATUM',SPECIES))
  p2012<-mutate(p2012,SPECIES=ifelse(SPECIES=='PSEUDOKEPHYRION PSUEDOSPIRALE','PSEUDOKEPHYRION PSEUDOSPIRALE',SPECIES))
  p2012<-mutate(p2012,SPECIES=ifelse(SPECIES=='PSEUDOKEPHYRION PSUEDOSPIRALE','PSEUDOKEPHYRION PSEUDOSPIRALE',SPECIES))

nrow(p2012)  #885
nrow(select(p2012,TAXA_ID)%>%distinct()) #885

phytoSp2012<-filter(p2012,!is.na(GENUS))%>%
          mutate(species=ifelse(is.na(SPECIES),GENUS,SPECIES))%>%
              select(ALGAL_GROUP,PHYLUM,CLASS,ORDER,FAMILY,GENUS,species)%>%
              distinct()%>%arrange(PHYLUM,GENUS)%>%
                mutate_each(funs(tolower))%>%mutate_each(funs(capitalize))%>%
                group_by(GENUS) %>%
                mutate(rank=row_number(species))%>%
                spread(rank,species)%>%
                  mutate(species=ifelse(is.na(`1`),NA,paste(`1`,`2`,`3`,`4`,`5`,`6`,`7`,`8`,`9`,
                          `10`,`11`,`12`,`13`,`14`,`15`,`16`,`17`,`18`,`19`,`20`,`21`,`22`,`23`,`24`,sep='; ')))%>%
                    mutate(species=gsub('NA; ','',species))%>%
                      mutate(species=gsub('; NA','',species))%>%
                        select(ALGAL_GROUP,PHYLUM,CLASS,ORDER,FAMILY,GENUS,sp2012=species)%>%ungroup()
                          
names(phytoSp2012)<-tolower(names(phytoSp2012))

#View(phytoSp2012)
```

* create a master taxa list 
* join phytoSp2007 and phytoSp2012 to create phytoSp
* add genus_id as rownumber
* save itis file as 'phytoItis.csv'

```{r master, include=FALSE, echo=FALSE}
#join the 2007 and 2012 taxa lists
p07<-phytoSp2007
p07<-rename(p07,phylum07=phylum,algal_group07=algal_group,order07=order,family07=family)

phytoSp<-full_join(phytoSp2012,p07)

phytoSp<-select(phytoSp,algal_group12=algal_group,algal_group07,phylum12=phylum,phylum07,class12=class,order12=order,order07,family12=family,family07,genus,sp12=sp2012,sp07=sp2007)

```

### create final taxonomy
* load the algaebase classification by genus ('data/workfiles/nla_algaebase_taxonomy.csv')
* join to the 2007 and 2012 taxononmy
* select fields to keep
* write 'output/nla_phyto_taxonomy.csv'


```{r betty, include=FALSE, echo=FALSE}
ab<-read.csv('data/workfiles/nla_algaebase_taxonomy.csv')
   names(ab)

nrow(ab) #362
nrow(phytoSp) #362

keep<-select(phytoSp,genus,sp07,sp12)

phytosp<-full_join(keep,ab)%>%select(phylum,subphylum,class,subclass,order,family,genus,genus_new,authority,sp07,sp12,flag,note,link)
nrow(phytosp) #362

write.csv(phytosp,'output/nla_phyto_taxonomy.csv',row.names = FALSE)
```
  
##QA/QC

* make sure all genera in the count data match observation in phytosp; all good

```{r qaqc, include=FALSE, echo=FALSE}
p2007<-read1('data/raw_data/','nla2007_phyto_count.csv');nrow(p2007) #31971
p2012<-read1('data/raw_data/','nla2012_phyto_count.csv');nrow(p2012) #38627
  p2012<-mutate(p2012,genus=ifelse(is.na(GENUS),NA,paste(toupper(substr(GENUS,1,1)),substr(GENUS,2,111),sep="")))


t2007<-left_join(p2007,phytosp,by=c("GENUS"="genus"));nrow(t2007) #31971
    filter(t2007,is.na(phylum))%>%select(GENUS)%>%distinct()
    
    
t2012<-left_join(p2012,phytosp,by="genus");nrow(t2012) #38627
    filter(t2012,is.na(phylum))%>%select(GENUS)%>%distinct()
```

    